```{r include = FALSE}
library(xtable)
printTable <- function(x, truncateNums = TRUE) {
  x <- xtable(x)
  if(truncateNums) {
    digits(x) <- c(0)
  }
  print(x, type = 'html', include.rowname = FALSE)
}
```

_Lattice_ vs _Ggplot2_ and _aggregate_ vs _plyr_
==========================================

Goal: to get a little more familiar with the awesome _ggplot2_ and _plyr_ functions by mimicking some results that were obtained with the older _lattice_ and _aggregate_

[Link to original tutorial](http://www.ugrad.stat.ubc.ca/~stat540/seminars/seminar04_compileNotebook-dataAggregation-twoGroupComparison.html)

#### Data preparation
Load the required libraries and data
```{r results='hide', message=FALSE, warning=FALSE}
library(lattice)
library(plyr)
library(ggplot2)
prDat <- read.table("GSE4051_data.tsv")
prDes <- readRDS("GSE4051_design.rds")
```

#### Single gene exploration

Extract data for one gene
```{r}
set.seed(987)
theGene <- sample(1:nrow(prDat), 1)
pDat <- data.frame(prDes, gExp = unlist(prDat[theGene, ]))
```

Explore!  
What are sample means in wildtype/knockout? First using _aggregate_, then using _plyr_
```
aggregate(gExp ~ gType, pDat, FUN = mean)
```
```{r results='asis', echo=F}
printTable(aggregate(gExp ~ gType, pDat, FUN = mean), truncateNums = FALSE)
```
```
ddply(pDat, ~ gType, summarize, gExp = mean(gExp))
```
```{r results='asis', echo=F}
printTable(ddply(pDat, ~ gType, summarize, gExp = mean(gExp)), truncateNums = FALSE)
```

Make sure the two actually returned identical results
```{r tidy=F}
identical(aggregate(gExp ~ gType, pDat, FUN = mean),
          ddply(pDat, ~ gType, summarize, gExp = mean(gExp)))
```

Plot!  
Strip plot of just the one gene, knockout vs wildtype. First using _lattice_, then using _ggplot2_
```{r fig.width=5, fig.height=3}
stripplot(gType ~ gExp, pDat)
ggplot(pDat, aes(x = gExp, y = gType)) + geom_point()
```

#### Multiple genes

Load in the dataset
```{r}
kDat <- readRDS("GSE4051_MINI.rds")
```

Explore!  
Average expression of eggBomb over developmental stages, first using _aggregate_, then using _plyr_
```
aggregate(eggBomb ~ devStage, kDat, FUN = mean)
```
```{r results='asis', echo=F}
printTable(aggregate(eggBomb ~ devStage, kDat, FUN = mean), truncateNums = FALSE)
```
```
ddply(kDat, ~ devStage, summarize, exp = mean(eggBomb))
```
```{r results='asis', echo=F}
printTable(ddply(kDat, ~ devStage, summarize, exp = mean(eggBomb)), truncateNums = FALSE)
```

Same thing, but now aggregate based on dev stage AND genotype
```
aggregate(eggBomb ~ gType * devStage, kDat, FUN = mean)
```
```{r results='asis', echo=F}
printTable(aggregate(eggBomb ~ gType * devStage, kDat, FUN = mean), truncateNums = FALSE)
```
```
ddply(kDat, .(gType, devStage), summarize, exp = mean(eggBomb))
```
```{r results='asis', echo=F}
printTable(ddply(kDat, .(gType, devStage), summarize, exp = mean(eggBomb)), truncateNums = FALSE)
```

#### Multigene plotting
Grab 6 genes: 3 interesting, 3 boring
```{r tidy=F}
keepGenes <- c("1431708_a_at", "1424336_at", "1454696_at",
               "1416119_at", "1432141_x_at", "1429226_at")
miniDat <- subset(prDat, rownames(prDat) %in% keepGenes)
miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))),
                      gene = factor(rep(rownames(miniDat), each = ncol(miniDat)),
                                    levels = keepGenes))
miniDat <- suppressWarnings(data.frame(prDes, miniDat))
```

Plot!  
Strip plot of the expression vs genotype, one plot per gene. First using _lattice_, then using _ggplot2_
```{r fig.width=8, fig.height=6, tidy=F}
stripplot(gType ~ gExp | gene, miniDat,
          scales = list(x = list(relation = "free")),
          group = gType, auto.key = TRUE)
ggplot(miniDat, aes(x = gExp, y = gType, color = gType)) +
  facet_wrap(~ gene, scales="free_x") +
  geom_point(alpha = 0.7) +
  theme(panel.grid.major.x = element_blank())
```
